Using Local PCA to Investigate Population Structure Along the Genome

### 2016-03-30 writeup/local_pca_paper.tex
Update local_pca_paper.tex

\section{Introduction}

The kinship matrix goes back to almost Mendel. It contains the kinship coefficient for pairwise individuals, which defines the genetic relatedness between individuals and could be estimated from genome-wide covariances of genotype markers. Populations are often structured in some way while there are systematic genetic variation. Linear Mixed models make use of the kinship matrix and can correct for population structure in GWAS. \citep{astle2009population,sul2013mixed}
About 37 years ago, \citet{menozzi1978synthetic} first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
PCA results are derived from the covariance matrix of genotype matrix, which could be used to estimate kinship matrix and describes information about population structure, demographic history, pedigree tree and so on.
The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{menozzi1978synthetic,novembre2008interpreting,mcvean2009genealogical}. Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. Plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent. The closeness of two samples in PC plots represents the similarity of their genetic information. Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. It is worth mentioning that this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns. \citep{astle2009population}
It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \citep{price2006principal}. 
However, for different part of genome, they have different genetic features. 
First, Locally everything is treelike and each site of DNA may have different gene tree. 
The covariance matrix of genotype matrix averages those gene trees. 
For a DNA region, if individuals have SNPs closer in gene tree, they tends to be more close in PC projections for that region. 
Different DNA segments may have different gene tree for their SNPs and would result in different population structure for those segments. 
Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
Third, if chromosome inversion exists, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints, thus results in different genome structure and population structure. 
Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure.

\plr{(This is starting to be methods/results: less detail?)}
Therefore, we cut the chromosomes into windows with hundreds/thousands of SNPs and apply PCA locally on each window. In this project, we used SNP data for human, \textit{Medicago truncatula}, and whole genome sequencing data for \textit{Drosophila}.

Based on the principal components, we can estimate the similarity of population structure contained in each genome window. To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \citep{borg2005modern}. To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, heterochromatin chunk and gene density along the genome. For different species, we got different explanation for the variation of population structure along the genome.

>>>>

\section{Introduction}
The kinship matrix contains the kinship coefficient for pairwise individuals. 
Kinship coefficient defines the genetic relatedness between individuals.
It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
Populations are often structured in some way while there are systematic genetic variation between populations.

About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
PCA results are derived from the covariance matrix of genotype matrix. 
The results of PCA can be directly related to the underlying genealogical history of the samples, 
such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

PCA is often used in genome-wide association studies (GWAS) for stratification correction \citep{price2006principal}. 
However, different parts of genome have different genetic features. 
First, each site of DNA may have different gene tree. 
The covariance matrix of genotype matrix averages those gene trees. 
For a genomic region, if individuals have alleles closer in gene tree, they tends to be more close in PC projections for that region. 
Different DNA segments may have different gene tree and therefore different population structure for those segments. 
Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
Selective sweeps cause local recent ancestry or short trees.
Balancing selection causes deep trees.
Background selection causes shallow ones.
Third, if a chromosome inversion is polymorphic in the sample, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints. Recombination suppression across inversions thus results in different genome structure and population structure. 
Other effects, like noise, introgression might also influence population structure.

Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure, and could possibly lead to more powerful methods for GWAS.

To investigate this, we cut each chromosome into windows (with hundreds to thousands of SNPs in each), applied PCA to each window, and visualized how population structure, as summarized by PCA, varies along windows.

In this project, we used SNP data for human, \textit{Medicago truncatula}, and whole genome sequencing data for \textit{Drosophila}.
Based on the principal components, we can estimate the similarity of population structure contained in each genome window. 
To quantify similarity of population structure between windows, we constructed for each window an approximate, scaled covariance matrix based on the first two PCs measured the pairwise Euclidean distance between those matrices.
We use multidimensional scaling to visualize the relationships between windows, which reduces the pairwise distance matrix to lower dimension while preserving the distance information between windows as well as possible \citep{borg2005modern}. 
To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, and heterochromatin and gene density along the genome. Each species showed distinct patterns, reflecting differences in their biology.

Other methods for visualizing population structure are like STRUCTURE, \citep{pritchard2000inference,falush2003inference, falush2007inference,hubisz2009inferring} model-based approach, \citep{yang2012modelbased} maps of heterozygosity \citep{ramachandran2005support}. The term ``local PCA" is also used in some other region for different meanings. \citep{manjon2013diffusion,kambhatla1997dimension,weingessel2000local}
